Current state and prospects of R-packages for the design of experiments

### 2022-12-14 paper/arxiv/arxiv.tex
reference  setups for specific outlets to accommodate differences in non-numbered/numbered sections and output types (PDF/HTML)

\usepackage{colortbl}
\usepackage{bera}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage[utf8]{inputenc}
\def\tightlist{}
\newcommand{\ctv}[1]{\href{http://CRAN.R-project.org/view=#1}{\emph{#1}}}

@@

\begin{abstract}
Re-running an experiment is generally costly and in some cases
impossible due to limited resources, so the design of an experiment
plays a critical role in increasing the quality of experimental data. In
this paper we describe the current state of the R-packages for the
design of experiments through an exploratory data analysis of package
downloads, package metadata, and the comparison of characteristics with
other topics. We observe that experimental designs in practice appear to
be sufficiently manufactured by a small number of packages and the
development of experimental designs often occur in silos. We discuss
also the interface designs of widely utilised R packages in the field of
experimental design and discuss its future prospects to advance the
field in practice.
\end{abstract}

@@

The critical role of data collection is well captured in the expression
``garbage in, garbage out'' -- in other words, if the collected data is
rubbish then no analysis, however complex it may be, can make something
out of it. A carefully crafted data collection scheme is therefore
critical to optimise the information from data. The field of
experimental designs is specifically devoted to planning the collection
of experimental data, largely based on the founding principles by Fisher
(1935) or an optimisation framework like those described in Pukelsheim
(2006). These experimental designs are often constructed with the aid of
a statistical software, such as R (R Core Team 2021), Python (Rossum
1995), SAS (SAS Institute 1985), and so on, so the usage of experimental
design software can inform us about some aspects of experimental designs
in practice.

Methods for data collection can be dichotomised by the type of data
collected -- namely, experimental or observational -- or alternatively,
categorised as experimental design (including quasi-experimental design)
or survey design. This dichotomisation, to a great extent, is seen in
the \href{https://cran.r-project.org/web/views/}{Comprehensive R Archive
Network (CRAN) task views} (a volunteer maintained list of R-packages by
topic) where R-packages for experimental design are in
\href{http://CRAN.R-project.org/view=ExperimentalDesign}{\emph{ExperimentalDesign}}
task view and R-packages for survey designs are in
\href{http://CRAN.R-project.org/view=OfficialStatistics}{\emph{OfficialStatistics}}
task view. The full list of available topics can be see in Table S1 in
the Supplementary Materials. A subset of experimental designs are
segregated into
\href{http://CRAN.R-project.org/view=ClinicalTrials}{\emph{ClinicalTrials}}
task view, where the focus is on clinical trials with primary interest
in sample size calculations. This paper focuses on the packages in
\href{http://CRAN.R-project.org/view=ExperimentalDesign}{\emph{ExperimentalDesign}}
task view, henceforth referred to as ``DoE packages''.

In
\href{http://CRAN.R-project.org/view=ExperimentalDesign}{\emph{ExperimentalDesign}}
task view, there are 114 R packages for experimental design and analysis
of data from experiments. The sheer quantity and variation of
experimental designs in the R-packages are arguably unmatched with any
other programming languages, e.g.~in Python, only a handful of packages
that generate design of experiment exist (namely \texttt{pyDOE},
\texttt{pyDOE2}, \texttt{dexpy}, \texttt{experimenter} and
\texttt{GPdoemd}) with limited type of designs. Thus, the study of DoE
packages, based on quantitative and qualitative data, can give us an
objective view to the state of the current experimental designs in
practice.

A utility of software can also be described by its design to facilitate
clear expression and interpretation of the desired experimental design.
Certain programming language design can hinder or discourage development
of a reliable program (Wasserman 1975). The immense popularity of
\texttt{tidyverse} (a collection of R-packages for various stages of
data analysis that place enormous emphasis on the interface design by
Wickham et al. 2019) is a testament to the impact an interface design
can have in practice. The practice of experimental design could be
advanced by adopting similar interface design principles across DoE
packages.

The paper is organised as follows. Section \ref{data} briefly describes
the data source used for the analysis in Section \ref{eda}; Section
\ref{eda} presents some insights into the state of the current DoE
packages by the exploratory data analysis of package download data, text
descriptions and comparisons with other CRAN task views; Section
\ref{design} discusses the interface designs of widely used DoE
packages, and we conclude with a discussion in Section \ref{discussion}
of future prospects in the software development of experimental designs.

@@

The Comprehensive R Archive Network (CRAN) is a network of servers
located across the world that store mirrored versions of R and
R-packages. The most popular network is the RStudio mirror (the default
server for those that use the RStudio IDE). The RStudio mirror is also
the only server that provides a comprehensive daily download logs of R
and R-packages since October 2012. The summary data can be easily
accessed with the \texttt{cranlogs} package (Csárdi 2019). This paper
uses the data from the beginning of 2013 to end of 2021 (a total of 9
years) for the packages in the CRAN task views.

@@

All CRAN packages have a title, description, package connections
(suggests, depends and imports) and other meta-information in the
DESCRIPTION file. We use the text data from the title and description
(as accessed in 2022-05-23) in Section \ref{topics}.

@@

CRAN task views are volunteer maintained list of R-packages on CRAN
relevant to the corresponding topic. There are a total 37 CRAN task
views. Table S1 in the Supplementary Materials list the full available
topics from the \texttt{ctv} package (Zeileis 2005). The list of
packages in each CRAN task views (as of 2022-05-25) are used to contrast
the characteristics of DoE packages in Sections \ref{popular} and
\ref{silo}.

\hypertarget{eda}{%
\section{Explorative data analysis}\label{eda}}

In this section, we derive some conjecture based on the analysis of the
data described in Section \ref{data}. All results presented are from the
exploratory data analysis of observational data, consequently, all
interpretations are somewhat speculative and may not be indicative of
the true state of the field of experimental design. In particular, any
analysis over time are confounded by the fact that the nature of users
and package management have also changed over the years. It should be
noted that some DoE packages may have been archived or removed from the
task view over the years so any cross-sectional analysis presented may
not reflect the set of all DoE packages at that particular time period
(although we assume such incidences are low).

A subset of DoE packages is not primarily about design of experiments
but about the analysis of experimental data. A complete delineation of
these packages is difficult as there is almost always at least one
function that can aid decisions or constructions of experimental designs
(and any categorisation is prone to our subjective bias) so we opted not
to remove any DoE packages in the analysis.

\hypertarget{popular}{%
\subsection{Small, but diverse, set of packages are sufficient for most
experimental designs in practice}\label{popular}}

There have been at least 50 DoE packages since 2013 but most of the
downloads are concentrated in just a handful of packages. For example,
Figure \ref{fig:plot-lorenz} shows a Lorenz curve (Lorenz 1905) for the
total package downloads in 2021 of 113 DoE packages (first released
prior to 2021); we can see from Figure \ref{fig:plot-lorenz} that bottom
90\% of DoE packages (in terms of total download count in 2021) only
share about 30\% of total downloads across all DoE packages -- in
another words, 70\% of the total downloads are due to 11 packages (10\%
of the DoE packages).

@@

\caption{Lorenz curve of the total download count for DoE packages in 2021. The red line corresponds to the line of perfect equality. The green region shows the area under the Lorenz curve and the red region shows the area of gap in equality.}\label{fig:plot-lorenz}
\end{figure}

If we consider package downloads as a measure of ``wealth'', then we can
consider using the Gini index (Gini 1921) as a measure of download
inequality across packages. The ratio of the red region over the total
colored regions in Figure \ref{fig:plot-lorenz} corresponds to the Gini
index for 2021. A Gini index of 0\% indicates equality in downloads
across packages while a value of 100\% indicates maximal inequality (all
downloads are due to one package). In Figure \ref{fig:download-share},
we see that the distributions of the package downloads each year have a
heavy right tail with the Gini index ranging from 42.6\% to 74.1\%
across the years 2013 to 2021, indicating that there is a high level of
inequality of package downloads, particularly with more pronounced
inequality in the last 6 years.

@@

\caption{The distribution of the number of downloads for each DoE package by year. Packages were removed in any year if it was released in that year or later so that each download count is for the full year. The label on the bottom of the plot shows the Gini index for downloads and the number of packages with full year of download count in the corresponding year. In the last 6 years, the Gini index is consistent above 60\% each year indicating that most downloads are due to a relatively small number of packages. }\label{fig:download-share}
\end{figure}

The increase in the number of packages that are not highly downloaded
may mean that \textbf{\emph{there are more packages to construct niche
experimental designs}}. Some examples of these packages include
\texttt{qtlDesign}, \texttt{PwrGSD} and \texttt{Crossover} made for QTL
experiments, group sequential designs and crossover trials,
respectively. These packages would naturally have a smaller number of
potential users. Counterfactual to this, the increase could be due to
other external factors, such as an increase in the number of skilled
developers (and thus more pacakge contributions), a change in CRAN
policy or management to add packages (either to CRAN and/or task view),
and/or that new packages are still yet to amass users. While there is an
argument that low download counts are due to the low utility and/or
quality of the packages, packages in CRAN task views are selected by
expert maintainers, thus we can reasonably assume that any package
listed in CRAN task view are of a decent utility and quality.

If the downloads are reflective of the experimental designs used in
practice, \textbf{\emph{small set of packages appear to be sufficient
for most to construct the full set of designs of experiments needed in
practice}}. Packages of course evolve and the top downloaded packages
have had regular updates that may have broadened its scope from its
previous releases.

While in absolute terms the Gini index is high for
\href{http://CRAN.R-project.org/view=ExperimentalDesign}{\emph{ExperimentalDesign}}
task view (42.6\% to 74.1\%), the inequality is not as severe as other
CRAN task views as shown in Figure \ref{fig:fig-gini-all-ctvs}. We can
see in Figure \ref{fig:fig-gini-all-ctvs} that the Gini index is
generally increasing for DoE packages (as is generally the case for
other CRAN task views as shown in Figure S1 in the Supplementary
Materials) but most other CRAN task views have a Gini index of over
75\%. This suggests that other CRAN task views may have dominant
standards and in comparison to other topics, there are more
\textbf{\emph{diverse approaches to designing experiments}} and thus no
single DoE package is dominant. This observation however doesn't take
into account other approaches to generating experimental designs, like
the proprietary software, CycDesignN (Whittaker, Williams, and John
2022) and SAS (SAS Institute 1985), which may be widely used.

@@

\caption{The points show the Gini index of the download counts by year for ExperimentalDesign task view with the color showing the number of packages. The grey line shows the distribution of the Gini index across years for all other CRAN task views. See Figure S1 in the Supplementary Material for the line graph for the Gini index across years by each CRAN task view.}\label{fig:fig-gini-all-ctvs}
\end{figure}

\hypertarget{ranking}{%
\subsection{The field of experimental design is
slow-changing}\label{ranking}}

We can see in Figure \ref{fig:rank-over-time} that most of the top 10
ranking packages have been in the top 10 for the last 9 years with
\texttt{lhs} steadily climbing up the ranks in the last few years. It
should be noted that the download of one package can prompt a download
of another package; the most notable dependency network is
\texttt{AlgDesign} and \texttt{agricolae}, where the former is an import
for the latter. The full dependency network within the DoE packages is
shown later in Figure @ref(fig:plot-doe-network).

@@

\caption{The plot shows the rank of top 10 downloaded packages by year. Packages that do not appear in top 10 for at least two periods are omitted from the plot. Most packages are consistently in the top 10 for the period shown.}\label{fig:rank-over-time}
\end{figure}

Figure \ref{fig:release-date-vs-download} shows a moderate negative
correlation between the first release date and the (log of) total
download counts of DoE packages in any given year from 2013 to 2021.
This suggest that in general, a package released earlier are more likely
to be used today (possibly for legacy reasons or the general inertia to
adopt new packages). We can also see in Figure
\ref{fig:release-date-vs-download} the most downloaded packages were
generally released in 2004 to 2010.

@@

\caption{The above figure shows the total download (in log scale) of a package in the corresponding year against the first release date of the package. The blue line corresponds to the least squares fit of a simple linear regression model. The label in the right-hand upper corner shows the sample correlation coefficient between the first release date and the log (with base 10) of the total download count. The high-leverage point on the far left belongs to `conf.design`, authored by one of the earlier contributors to R.}\label{fig:release-date-vs-download}
\end{figure}

The consistency in the top 10 ranking packages (Figure
\ref{fig:rank-over-time}) and the fact that most downloaded DoE packages
are first released more than 10 years ago (Figure
\ref{fig:release-date-vs-download}) indicates that either the existing
packages filled the needs of mass in practice or no new packages were
compelling for many to switch their practice. We do however also see
that the the top downloaded packages generally have more updates (see
Figure \ref{fig:release-date-vs-download}) so it is possible that the
packages have improved or broadened the scope of its usage.

\hypertarget{topics}{%
\subsection{Optimal designs are of interest}\label{topics}}

Figure \ref{fig:wordcloud-over-time} shows some common purposes of DoE
packages, based on bigrams in the package title and description. We show
only the bigrams as unigram was not insightful and there were not many
trigrams common across packages. In counting the bigrams, we processed
the text data as follows:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  We standardised the words to lower case and removed pluralization.
\item
  Multiple mentions of the same bigram within a package was counted as
  one (e.g.~\texttt{AlgDesign} mentions ``experimental design'' four
  times in the title and the description but this is counted as one).
\item
  Bigrams that consist the stop words (provided in
  \texttt{tidytext::stop\_words} in addition to other words we deemed
  irrelevant, e.g.~``provide'', ``e.g.'', ``calculate'' and so on; full
  list is shown in the code provided in the link under Section
  \ref{pkgs}) are removed.
\end{enumerate}

Not surprisingly, the bigram ``experimental design'' was the most
common. More interestingly, ``optimal design'' and ``sequential design''
appeared across different packages (indicated by the size of the word in
Figure \ref{fig:wordcloud-over-time}), and the bigrams ``latin
hypercube'' and ``computer experiment'' are used across a few packages
that are downloaded frequently (indicated by the color of the word in
Figure \ref{fig:wordcloud-over-time}). Sequential design, Latin
hypercube sampling and computer experiments (which generally include
space filling designs like Latin hypercube sampling) are designs that
generally operate by optimising a user-selected criterion and can be
classified as optimal designs.

@@

\caption{The above figure shows the word cloud of bigrams from the title and descriptions of the DoE packages. The size shows how often the bigram appears across the DoE packages and the color are relative to the total download count in 2021 of the packages that contain the bigram.}\label{fig:wordcloud-over-time}
\end{figure}

@@

\hypertarget{silo}{%
\subsection{Development of experimental designs occur in
silos}\label{silo}}

Figure \ref{fig:ctv-summ-plot} shows that the
\href{http://CRAN.R-project.org/view=ExperimentalDesign}{\emph{ExperimentalDesign}}
task view has the lowest average number of contributors among all 37
CRAN task views. In addition, we can also see in Figure
\ref{fig:ctv-summ-plot} that the
\href{http://CRAN.R-project.org/view=ExperimentalDesign}{\emph{ExperimentalDesign}}
task view has one of the least intra-connectivity (the percentage of
packages that make use of other packages within the same task view). The
full connection of DoE packages with other DoE packages is shown in
Figure \ref{fig:plot-doe-network}. These observations suggest that the
\textbf{\emph{field of experimental design is one of the least
collaborative field}} and package development generally occur in silos.

@@

\caption{The above figure is a scatterplot of the intra-connectivity (the percentage of packages that depends, suggest or imports at least one other package within the same task view) and the average number of contributors for each CRAN task view. A low intra-connectivity suggests that development within the topic mostly occur in silos whilst high  intra-connectivity suggests that there are more interactions within the topic. The color shows the number of packages, the size of the point corresponds to the total number of contributors, and the text labels show the CRAN task view name.  The label of ExperimentalDesign task view is colored red. The task views in the bottom-left corner are topics that are more indicative of contributors working in silos. The actual numerical values are show in Table S1 in the Supplementary Material.}\label{fig:ctv-summ-plot}
\end{figure}

@@

\caption{Package connections (depends, suggests and imports) within the DoE packages. The direction of the arrow shows the connection of packags where the package on the tail of the arrow is a depedency, suggestion or import for the package on the head of the arrow.  DoE packages that do not depend, suggest or import another DoE package is not shown.}\label{fig:plot-doe-network}
\end{figure}

@@

In software design, there are two interface designs to consider: user
interface (UI) and application programming interface (API). UI is
concerned with the interaction of the software by the user, while the
API is concerned with how different programs interact and is
predominately of the interest to the developer. The UI design is an
abstraction to specifying the desired experimental design and its
choices enable how a user expresses the specification of the
experimental design. The API design aids in other developers to leverage
existing systems.

In this section, we discuss these interface designs broadly, with
examples illustrated from the top downloaded packages (shown in Figure
\ref{fig:rank-over-time}). We exclude \texttt{ez} and
\texttt{DiceKriging} from the discussion as the former is predominately
visualisation of experimental data and the latter is about the analysis
of computer experiments in addition to belonging to the same suite of
packages as \texttt{DiceDesign} (Dupuy, Helbert, and Franco 2015).

>>>>

\usepackage{colortbl}
\usepackage{bera}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage[utf8]{inputenc}
\def\tightlist{}
\newcommand{\ctv}[1]{\href{http://CRAN.R-project.org/view=#1}{\emph{#1}}}
\newcommand{\CRANpkg}[1]{\href{https://cran.r-project.org/web/packages/#1/index.html}{\texttt{#1}}}

@@

\begin{abstract}
Re-running an experiment is generally costly and, in some cases,
impossible due to limited resources; therefore, the design of an
experiment plays a critical role in increasing the quality of
experimental data. In this paper, we describe the current state of
R-packages for the design of experiments through an exploratory data
analysis of package downloads, package metadata, and a comparison of
characteristics with other topics. We observed that experimental designs
in practice appear to be sufficiently manufactured by a small number of
packages, and the development of experimental designs often occurs in
silos. We also discuss the interface designs of widely utilized R
packages in the field of experimental design and discuss their future
prospects for advancing the field in practice.
\end{abstract}

@@

The critical role of data collection is well captured in the expression
``garbage in, garbage out'' -- in other words, if the collected data are
rubbish, then no analysis, however complex it may be, can make something
out of it. Therefore, a carefully crafted data-collection scheme is
critical for optimizing the information from the data. The field of
experimental design is specifically devoted to planning the collection
of experimental data, largely based on the founding principles of Fisher
(1935) or an optimization framework like those described in Pukelsheim
(2006). These experimental designs are often constructed with the aid of
statistical software such as R (R Core Team 2021), Python (Rossum 1995),
and SAS (SAS Institute 1985); thus the use of experimental design
software can inform us about some aspects of experimental designs in
practice.

Methods for data collection can be dichotomized by the type of data
collected -- namely, experimental or observational -- or alternatively,
categorized as experimental design (including quasi-experimental design)
or survey design. This dichotomization, to a great extent, is seen in
the \href{https://cran.r-project.org/web/views/}{Comprehensive R Archive
Network (CRAN) task views} (a volunteer maintained list of R-packages by
topic) where R-packages for experimental design are in
\href{http://CRAN.R-project.org/view=ExperimentalDesign}{\emph{ExperimentalDesign}}
task view and R-packages for survey designs are in
\href{http://CRAN.R-project.org/view=OfficialStatistics}{\emph{OfficialStatistics}}
task view. A full list of available topics is provided in Table S1 in
the Supplementary Materials. A subset of experimental designs is
segregated into the
\href{http://CRAN.R-project.org/view=ClinicalTrials}{\emph{ClinicalTrials}}
task view, where the focus is on clinical trials with primary interest
in sample size calculations. This paper focuses on packages in
\href{http://CRAN.R-project.org/view=ExperimentalDesign}{\emph{ExperimentalDesign}}
task view, henceforth referred to as ``DoE packages''.

From the
\href{http://CRAN.R-project.org/view=ExperimentalDesign}{\emph{ExperimentalDesign}}
task view, there are 105 R packages for the experimental design and
analysis of data from experiments. The sheer quantity and variation of
experimental designs in the R-packages are arguably unmatched with any
other programming languages, for example, in Python, only a handful of
packages that generate design of experiment exist (namely
\texttt{pyDOE}, \texttt{pyDOE2}, \texttt{dexpy}, \texttt{experimenter},
and \texttt{GPdoemd}) with a limited type of design. Thus, the study of
DoE packages, based on quantitative and qualitative data, can provide an
objective view of the state of current experimental designs in practice.

The utility of the software can also be described by its design to
facilitate the clear expression and interpretation of the desired
experimental design. Certain programming language designs can hinder or
discourage the development of reliable programs (Wasserman 1975). The
immense popularity of
\href{https://cran.r-project.org/web/packages/tidyverse/index.html}{\texttt{tidyverse}}
(a collection of R-packages for various stages of data analysis that
places enormous emphasis on the interface design by Wickham et al. 2019)
is a testament to the impact that an interface design can have in
practice. The practice of experimental design can be advanced by
adopting similar interface design principles across the DoE packages.

The remainder of this paper is organized as follows. Section \ref{data}
briefly describes the data source used for the analysis; Section
\ref{eda} presents some insights into the state of the current DoE
packages by the exploratory data analysis of package download data, text
descriptions and comparisons with other CRAN task views; Section
\ref{design} discusses the interface designs of widely used DoE
packages, and we conclude with a discussion in Section \ref{discussion}
of future prospects in the software development of experimental designs.

@@

The Comprehensive R Archive Network (CRAN) is a network of servers
located across the world that stores mirrored versions of the R and R
packages. The most popular network is the RStudio mirror (the default
server for those that use the RStudio IDE). The RStudio mirror is also
the only server that provides a comprehensive daily download logs of R
and R packages since October 2012. The summary data can be easily
accessed using the
\href{https://cran.r-project.org/web/packages/cranlogs/index.html}{\texttt{cranlogs}}
package (Csárdi 2019). This paper uses the data from the beginning of
2013 to the end of 2021 (a total of nine years) for the packages in the
CRAN task views.

@@

All CRAN packages have a title, description, package connections
(suggests, depends, and imports of other packages), and other
meta-information in the DESCRIPTION file. We use text data from the
title and description (accessed in 2022-12-12).

@@

CRAN task views are volunteer-maintained lists of R-packages on CRAN
relevant to the corresponding topic. There were 39 CRAN task views in
total. Table S1 in the Supplementary Materials list the available topics
from the
\href{https://cran.r-project.org/web/packages/ctv/index.html}{\texttt{ctv}}
package (Zeileis 2005). The list of packages in each CRAN task view (as
of 2022-12-12) is used to contrast the characteristics of the DoE
packages.

\hypertarget{eda}{%
\section{Exploratory data analysis}\label{eda}}

In this section, we derive some conjectures based on an analysis of the
data described in Section \ref{data}. All results presented are from
exploratory data analysis of observational data, consequently, all
interpretations are somewhat speculative and may not be indicative of
the true state of the field of experimental design. In particular, any
analysis over time is confounded by the fact that the nature of users
and package management has changed over the years. It should be noted
that some DoE packages may have been archived or removed from the task
view over the years; therefore, any cross-sectional analysis presented
may not reflect the set of all DoE packages at that particular time
period (although we assume such incidences are low).

A subset of DoE packages is not primarily about the design of
experiments but about the analysis of experimental data. A complete
delineation of these packages is difficult, as there is almost always at
least one function that can aid decisions or constructions of
experimental designs (and any categorization is prone to our subjective
bias); therefore we opted not to remove any DoE packages in the
analysis.

\hypertarget{small-but-diverse-set-of-packages-are-sufficient-for-most-experimental-designs-in-practice}{%
\subsection{Small, but diverse, set of packages are sufficient for most
experimental designs in
practice}\label{small-but-diverse-set-of-packages-are-sufficient-for-most-experimental-designs-in-practice}}

There are at least 50 DoE packages since 2013 but most of the downloads
are concentrated in only a handful of packages. For example, Figure
\ref{fig:plot-lorenz} shows a Lorenz curve (Lorenz 1905) for the total
package downloads in 2021 for 102 DoE packages (first released prior to
2021). We can see from Figure \ref{fig:plot-lorenz} that the bottom 90\%
of DoE packages (in terms of total download count in 2021) only share
approximately 32\% of total downloads across all DoE packages; in other
words, 68\% of the total downloads are due to 10 packages (10\% of the
DoE packages).

@@

\caption{Lorenz curve of the total download count for DoE packages in 2021. The red line corresponds to the line of perfect equality. The yellow region shows the area under the Lorenz curve and the red region shows the area of the gap in equality.}\label{fig:plot-lorenz}
\end{figure}

If we consider package downloads as a measure of ``wealth'', then we can
consider using the Gini index (Gini 1921) as a measure of download
inequality across packages. The ratio of the red region to the total
colored regions in Figure \ref{fig:plot-lorenz} corresponds to the Gini
index for 2021. A Gini index of 0\% indicates equality in downloads
across packages whereas a value of 100\% indicates maximal inequality
(all downloads are due to one package). In Figure
\ref{fig:download-share}, we see that the distributions of the package
downloads each year have a heavy right tail with the Gini index ranging
from 32.7\% to 69.1\% across the years 2013 to 2021, indicating that
there is a high level of inequality in package downloads, particularly
with more pronounced inequality in the last six years.

@@

\caption{Distribution of number of downloads for DoE packages by year. Packages were removed in any year if they were released in that year or later so that each download count was for the full year. The label at the bottom of the plot shows the Gini index for downloads and the number of packages with a full download count in the corresponding year. In the last six years, the Gini index has consistently exceeded 60\%, indicating that most downloads are due to a relatively small number of packages. }\label{fig:download-share}
\end{figure}

An increase in the number of packages that are not highly downloaded may
mean that \textbf{\emph{there are more packages to construct niche
experimental designs}}. Some examples of these packages include
\href{https://cran.r-project.org/web/packages/qtlDesign/index.html}{\texttt{qtlDesign}},
\href{https://cran.r-project.org/web/packages/PwrGSD/index.html}{\texttt{PwrGSD}}
and
\href{https://cran.r-project.org/web/packages/Crossover/index.html}{\texttt{Crossover}}
made for QTL experiments, group sequential designs and crossover trials,
respectively. These packages would naturally have fewer potential users.
Counterfactual to this, the increase could be due to other external
factors, such as an increase in the number of skilled developers (and
thus more package contributions), a change in CRAN policy or management
to add packages (either to CRAN and/or task view), and/or the fact that
new packages are still yet to amass users. While there is an argument
that low download counts are due to the low utility and/or quality of
packages, packages in CRAN task views are selected by expert
maintainers. We can reasonably assume that any package listed in the
CRAN task view has an acceptable utility and quality.

If the downloads are reflective of the experimental designs used in
practice, \textbf{\emph{a small set of packages appears to be sufficient
for most users to construct the full set of designs of experiments they
need in practice}}. Packages of course evolve, and the top downloaded
packages have had regular updates that may have broadened their scope
from their previous releases.

While in absolute terms the Gini index is high for the
\href{http://CRAN.R-project.org/view=ExperimentalDesign}{\emph{ExperimentalDesign}}
task view (32.7\% to 69.1\%), the inequality is not as severe as in
other CRAN task views as shown in Figure \ref{fig:fig-gini-all-ctvs}. We
can see in Figure \ref{fig:fig-gini-all-ctvs} that the Gini index
generally increases over time for DoE packages (as is generally the case
for other CRAN task views as shown in Figure S1 in the Supplementary
Materials), but most other CRAN task views have a Gini index of over
75\%. This suggests that other CRAN task views may have dominant
standards, and in comparison to other topics, there are more
\textbf{\emph{diverse approaches to designing experiments}}, and thus,
no single DoE package is dominant. However, this observation doe not
consider other approaches to generate experimental designs, such as the
proprietary software, CycDesignN (Whittaker, Williams, and John 2022),
which may be widely used.

@@

\caption{The points show the Gini index of the download counts by year for the ExperimentalDesign task view with the color showing the number of packages. The gray lines show the line plots of the Gini index across years for all other CRAN task views. See Figure S1 in the Supplementary Material for the line graph of the Gini index across years for each CRAN task view.}\label{fig:fig-gini-all-ctvs}
\end{figure}

\hypertarget{the-field-of-experimental-design-is-slow-changing}{%
\subsection{The field of experimental design is
slow-changing}\label{the-field-of-experimental-design-is-slow-changing}}

We can see in Figure \ref{fig:rank-over-time} that most of the top 10
ranking packages have been in the top 10 for the last nine years with
\href{https://cran.r-project.org/web/packages/lhs/index.html}{\texttt{lhs}}
steadily climbing up the ranks in the last few years. It should be noted
that the download of one package can prompt the download of another
package; the most notable package connection is
\href{https://cran.r-project.org/web/packages/AlgDesign/index.html}{\texttt{AlgDesign}}
and
\href{https://cran.r-project.org/web/packages/agricolae/index.html}{\texttt{agricolae}},
where the former is an import for the latter. The full network of
package connections within the DoE packages is shown in Figure
\ref{fig:plot-doe-network}.

@@

\caption{The plot shows the rank of the top 10 packages downloaded by year. Packages that did not appear in the top 10 for at least two periods were omitted from the plot. Most packages are consistently in the top 10 for the period shown.}\label{fig:rank-over-time}
\end{figure}

Figure \ref{fig:release-date-vs-download} shows a moderate negative
correlation between the first release date and the (log of) total
download counts of DoE packages for any given year from 2013 to 2021.
This suggests that packages released earlier are more likely to be used
today (possibly for legacy reasons or the general inertia to adopt new
packages). We can also see in Figure \ref{fig:release-date-vs-download}
that most downloaded packages were released from 2004 to 2010.

@@

\caption{The above figure shows the total download (in log scale) of a package in the corresponding year against the first release date of the package. The blue line corresponds to the least-squares fit of the simple linear regression model. The label in the upper right-hand corner shows the sample correlation coefficient between the first release date and log (with base 10) of the total download count. The high leverage point on the far left belongs to `conf.design`, authored by one of the earlier contributors to R.}\label{fig:release-date-vs-download}
\end{figure}

The consistency in the top 10 ranking packages (Figure
\ref{fig:rank-over-time}) and the fact that most downloaded DoE packages
were first released more than 10 years ago (Figure
\ref{fig:release-date-vs-download}) indicate that either the existing
packages fulfilled the needs of mass in practice or no new packages were
compelling for many to switch their practice. However, we also see that
the top downloaded packages generally have more updates (see Figure
\ref{fig:release-date-vs-download}); therefore, it is possible that the
packages have improved or broadened the scope of their usage.

\hypertarget{optimal-designs-are-of-interest}{%
\subsection{Optimal designs are of
interest}\label{optimal-designs-are-of-interest}}

Figure \ref{fig:wordcloud-over-time} shows some of the common purposes
of DoE packages, based on bigrams in the package title and description.
We only show the bigrams as unigrams were not insightful, and there were
not many trigrams common across packages. To count the bigrams, we
processed the text data as follows:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  We standardized the words to the lower case and removed pluralization.
\item
  Multiple mentions of the same bigram within a package were counted as
  one (for example,
  \href{https://cran.r-project.org/web/packages/AlgDesign/index.html}{\texttt{AlgDesign}}
  mentions ``experimental design'' four times in the title and the
  description, but this is counted as one).
\item
  Bigrams consisting of stop words were removed. The stop words are
  sourced from the lexicons in \texttt{tidytext::stop\_words} in
  addition to other words we deemed irrelevant, e.g.~``provide'',
  ``e.g.'', ``calculate'' and so on -- full list is shown in the code
  provided in the link under Section \ref{pkgs}.
\end{enumerate}

Unsurprisingly, the bigram ``experimental design'' was the most common.
More interestingly, ``optimal design'' and ``sequential design''
appeared across different packages (indicated by the size of the word in
Figure \ref{fig:wordcloud-over-time}), and the bigrams ``latin
hypercube'' and ``computer experiment'' are used across a few packages
that are downloaded frequently (indicated by the color of the word in
Figure \ref{fig:wordcloud-over-time}). Sequential design, Latin
hypercube sampling and computer experiments (which generally include
space-filling designs such as Latin hypercube sampling) generally
operate by optimizing a user-selected criterion and can be classified as
optimal designs.

@@

\caption{The above figure shows the word cloud of bigrams from the title and descriptions of the DoE packages. The size shows how often the bigram appears across the DoE packages and the color is relative to the total download count in 2021 for the packages that contain the bigram.}\label{fig:wordcloud-over-time}
\end{figure}

@@

\hypertarget{development-of-experimental-designs-occur-in-silos}{%
\subsection{Development of experimental designs occur in
silos}\label{development-of-experimental-designs-occur-in-silos}}

Figure \ref{fig:ctv-summ-plot} shows that the
\href{http://CRAN.R-project.org/view=ExperimentalDesign}{\emph{ExperimentalDesign}}
task view has the lowest average number of contributors among all the 39
CRAN task views. In addition, we can also see in Figure
\ref{fig:ctv-summ-plot} that the
\href{http://CRAN.R-project.org/view=ExperimentalDesign}{\emph{ExperimentalDesign}}
task view has one of the least intra-connectivity (the percentage of
packages that make use of other packages within the same task view). The
full connection between DoE packages is shown in Figure
\ref{fig:plot-doe-network}. These observations suggest that
\textbf{\emph{experimental design is one of the least collaborative
fields}} and package development generally occurs in silos.

@@

\caption{The above figure is a scatterplot of intra-connectivity (the percentage of packages that depends, suggests or imports at least one other package within the same task view) and the average number of contributors for each CRAN task view. Low intra-connectivity suggests that development within the topic mostly occurs in silos, while high  intra-connectivity suggests that there are more interactions within the topic. The color shows the number of packages, the size of the point corresponds to the total number of contributors, and the text labels show the CRAN task view names.  The label of ExperimentalDesign task view is colored in red. The task views in the bottom-left corner are topics that are more indicative of contributors working in silos. The actual numerical values are listed in Table S1 in the Supplementary Material.}\label{fig:ctv-summ-plot}
\end{figure}

@@

\caption{Package connections (depends, suggests and imports) within DoE packages. The direction of the arrow shows the connection of packags where the package on the tail of the arrow is a dependency, suggestion or import for the package on the head of the arrow.  DoE packages that do not depend, suggest or import another DoE package are not shown.}\label{fig:plot-doe-network}
\end{figure}

@@

In software design, there are two interface designs to consider: user
interface (UI) and application programming interface (API). The UI is
concerned with the interaction of the software by the user, while the
API is concerned with how different programs interact and is
predominately of interest to the developer. The UI design is an
abstraction that specifies the desired experimental design, and its
choices enable how a user expresses the specification of an experimental
design. The API design aids other developers in leveraging existing
systems.

In this section, we discuss the interface designs of functions that
output an experimental design based on three broad areas: factorial,
recipe and augmenting designs. The discussion is exclusive to the top
downloaded packages (shown in Figure \ref{fig:rank-over-time}), with the
exception of
\href{https://cran.r-project.org/web/packages/ez/index.html}{\texttt{ez}}
and
\href{https://cran.r-project.org/web/packages/DiceKriging/index.html}{\texttt{DiceKriging}},
as the former is predominately visualization of experimental data and
the latter is about the analysis of computer experiments in addition to
belonging to the same suite of packages as
\href{https://cran.r-project.org/web/packages/DiceDesign/index.html}{\texttt{DiceDesign}}
(Dupuy, Helbert, and Franco 2015).
